---
title: "Explore timetables data"
output: html_document
---

Adam Bricknell, November 2021

```{r setup, include=TRUE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(dplyr )
library(purrr)
library(data.table)
library(ggplot2)
```

## Isle of Wight timetables

The XML files for the Isle of Wight buses have been extracted into 8 tables. We explore them and how they link together here. We also attempt to link them to AVL. 

```{r, echo=TRUE}
load("full_isle_of_wight.RData")

print(map(full_isle_of_wight, dim))
```


## Stop points

Gives you ID code, name and lat/long of bus stops. There are 981 bus stops on the Isle of Wight

```{r, echo=TRUE}
head(full_isle_of_wight$stop_points)
```



## Route sections

Individual 'links', or line segments, for each route. May enable plotting of the route. 

routelink = section of route (line segment). Links to 'RouteLinkRef' in journey_pattern_sections table

route_section_id = I *think* this corresponds to 'id' in Routes (there are the same number of codes). However the codes don't match up. 



```{r, echo=TRUE}
print(unique(full_isle_of_wight$all_route_sections_df$route_section_id))
head(full_isle_of_wight$all_route_sections_df)
```

Below are routes for one service (everything from one XML file). Looks like there is some noise in the coordinates of routes: perhaps it uses old AVL data or maybe something was missed in the processing. 

```{r, echo=TRUE}
full_isle_of_wight$all_route_sections_df %>% 
  filter(file_source == 'Vectis 26 09 2021_SER1.xml') %>%
  ggplot(aes(x = latitude, 
             y = longitude, 
             group = routelink, 
             color = route_section_id)) + 
    geom_line() +
    ggthemes::theme_few() +
    theme(axis.text.y = element_blank(),
          axis.text.x = element_blank())

```

Below we plot every link on the Isle of Wight by service. 

```{r, echo=TRUE}
full_isle_of_wight$all_route_sections_df %>% 
  mutate(ServiceCode = as.factor(ServiceCode)) %>%
  ggplot(aes(x = latitude, 
             y = longitude, 
             group = unique_service_routelink_id, 
             color = ServiceCode)) + 
    geom_line() +
    ggthemes::theme_few() +
    theme(axis.text.y = element_blank(),
          axis.text.x = element_blank())

```


Zooming in on several routelinks for one service

```{r, echo=TRUE}
full_isle_of_wight$all_route_sections_df %>% 
  filter(routelink %in% c('RL1', 'RL2', 'RL3', 'RL4', 'RL5', 'RL6', 'RL7'),
         ServiceCode == 'SER1') %>%
ggplot(aes(x = latitude, y = longitude, group = routelink, color = routelink)) + 
  geom_line() +
  ggthemes::theme_few() +
  theme(axis.text.y = element_blank(),
        axis.text.x = element_blank())

```


## Routes

Gives you unique route IDs

```{r, echo=TRUE}
head(full_isle_of_wight$routes)
```


## Journey patterns

Gives each section of journey between bus stops (From.StopPointRef and To.StopPointRef : link to stop_points table) for each unique journey (journey_id)

RouteLinkRef = matches to 'routelink' in all_route_sections_df table (say, perhaps, if you wanted to plot the routes between each stop)


```{r, echo=TRUE}
head(full_isle_of_wight$journey_pattern_sections)
```



## Operators

Operator details for each service. Can join to Service Details with ServiceCode


```{r, echo=TRUE}
head(full_isle_of_wight$operators)
```


## Service details

Operator details for each service

```{r, echo=TRUE}
head(full_isle_of_wight$service_details)
```

## Standard service

Shows ultimate destination at each link in the service. And whether inbound or outbound.

```{r, echo=TRUE}
head(full_isle_of_wight$standard_service)
```



## Vehicle journey and departure times

Includes vehicle type, ticket machine codes, journey code (VehicleJourneyCode), Garage used (GarageRef), Service code (ServiceRef), Journey pattern (links to journey_pattern_sections, however the lookup codes will need wrangling to remove the 'S' in JourneyPatternRef), VehicleJourneyCode and Departure time

There last 10 columns are sparsely populated: there aren't included for many routes

```{r, echo=TRUE}
head(full_isle_of_wight$vehicle_journeys_and_departure_times)
```

Below we see the distribution of departure times. 

```{r, echo=TRUE}
hist <- lubridate::parse_date_time(full_isle_of_wight$vehicle_journeys_and_departure_times$DepartureTime, orders = '%H:%M:%S') %>% 
  as.data.frame() %>%
  magrittr::set_colnames('time') %>%
  ggplot(aes(x = time, alpha = 0.6)) + 
    geom_histogram(fill='purple',colour = 'black', bins = 48) +
    ggthemes::theme_few() +
    theme(legend.position = 'none')

hist

```





## Does JourneyPatternRef link to journey_id?

Looks like they might (all pattern_refs are in journey_ids, but the reverse isnt true) However the lookup codes will need wrangling to remove the 'S' in JourneyPatternRef.

```{r, echo=TRUE}
journey_ids <- unique(full_isle_of_wight$journey_pattern_sections$journey_id) %>% substring(4)
pattern_refs <- unique(full_isle_of_wight$vehicle_journeys_and_departure_times$JourneyPatternRef) %>% substring(3)

print(pattern_refs %in% journey_ids)
print(journey_ids %in% pattern_refs) 

```







